BANTER for Beaked Whales: South Atlantic

### 2023-03-02 hawaii/beakerBanter_hawaii.Rmd
add epacific analysis
Added updated epacific analysis using CCES

---
title: "BANTER for Beaked Whales: Hawaii"
author: "Shannon Rankin"
date: "2022-02-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
saveRDS(banter_data_ici, file='hawaii_banter_data_ici.rds')
```

>>>>

---
title: "BANTER for Beaked Whales: Hawaii"
author: "Shannon Rankin"
date: "2023-02-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

banter_data_ici <- export_banter(data_ch1only, dropSpecies = c("BW", "possBW"), training=TRUE)
saveRDS(banter_data_ici, file='hawaii_banter_data_ici.rds')

#save update of Acoustic Study
saveRDS(data_ch1only, 'hawaii_study.rds')
```

### 2023-03-02 manuscript/manuscript.qmd
add epacific analysis
Added updated epacific analysis using CCES

```{r, hide=TRUE}
library(tidyverse)
library(knitr)
library(here)

@@

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods ([@macleod2018]). However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering2013a]. Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow2022]; [@baumann-pickering2014]; [@simonis2020]. Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serve to improve the efficiency and decrease the cost of analyzing large datasets.

@@

@@

@@

## Hawaii
The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al. 2018); species included Longman's beaked whale (n=121), Blainsville's beaked whale (n=518), Cross Seamount beaked whale (n=84) and Cuvier's beaked whale (n=201). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

>>>>

```{r, hide=TRUE}
library(tidyverse)
library(PAMpal)
library(knitr)
library(here)
here
```

@@

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod2018]. However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering2013a]. Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow2022]; [@baumann-pickering2014]; [@simonis2020]. Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serve to improve the efficiency and decrease the cost of analyzing large datasets.

@@

```{r, hide = TRUE}
setwd("..")
natlanticData<-readRDS("natlantic/natlantic_study.rds")
#Note: Species IDs not saved. Need to rerun? Not sure whats up.

@@

```{r, hide = TRUE}
setwd("..")
satlanticData<-readRDS("satlantic/satlantic_study.rds")
```

@@

## Hawaii

The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (Yano et al. 2018); species included Longman's beaked whale (n=121), Blainsville's beaked whale (n=518), Cross Seamount beaked whale (n=84) and Cuvier's beaked whale (n=201). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

### 2023-03-02 manuscript/manuscript.tex
add epacific analysis
Added updated epacific analysis using CCES

\title{Open-Source machine learning BANTER acoustic classification of a
cryptic echolocating species}
\author{Shannon Rankin \and Taiki Sakai \and Frederick Archer \and Jay
Barlow \and Danielle Cholewiak \and Annamaria DeAngelis \and Jennifer L.
K. McCullough \and Erin Oleson \and Melissa Soldevilla}
\date{2/6/23}

\begin{document}
\maketitle
\begin{abstract}
Passive acoustic monitoring is increasingly used for assessing
populations of marine mammals; however, analysis of large datasets is
limited by our ability to easily classify sounds detected.
Classification of beaked whale acoustic events, in particular, require
evaluation of multiple lines of evidence by expert analysts. Here we
present a highly automated approach to acoustic detection and
classification using supervised machine learning and open source
software methods. Data from four large scale surveys of beaked whales
(North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were
analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and
BANTER (hierarchical random forest classifier). Overall correct
classification results ranged from 86\% for the South Atlantic data to
99\% for the North Atlantic. Results for many species could likely be
improved with increased sample sizes and consideration of alternative
automated detectors. These methods provide a highly automated approach
to acoustic detection and classification using open source methods that
can be readily adopted for species and geographic regions.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Passive acoustic monitoring (PAM) from has proven to be a valuable tool
for studying populations of marine mammals ((Parijs et al. 2009));
however, the value of these studies depends on our ability to identify
the sources of the sounds we are monitoring. Historically, experienced
acousticians manually identify stereotyped sounds that could be reliably
attributed to a given species, based on their spectral or temporal
characteristics ((Baumann-Pickering et al. 2013a); (Bittle and Duncan
2013); (Rankin and Barlow 2005); (Soldevilla et al. 2008)). The dramatic
increase in recordings make it impossible for experienced acousticians
to manually annotate all data, and there has been an increasing need for
the development of automated classification routines that can provide
accurate determinations of the species responsible for the sounds.

Beaked whales are deep diving marine mammals found in offshore waters;
their long dive intervals and cryptic surfacing behavior make them
difficult to study using typical shipboard visual observation methods
((MacLeod 2018)). However, beaked whales make stereotyped echolocation
clicks that have temporal and spatial characteristics that can be used
to differentiate species ((Baumann-Pickering et al. 2013b)). Detection
of beaked whale signals in large datasets is greatly expanding our
understanding of their population structure and the potential impact of
human activities on these species ((Barlow et al. 2022);
(Baumann-Pickering et al. 2014); (Simonis et al. 2020)). Unfortunately,
manual classification of beaked whale echolocation clicks requires
analysis by trained observers using a number of visual aids to examine
the call characteristics. Development of automated classification
routines, if accurate, serve to improve the efficiency and decrease the
cost of analyzing large datasets.

BANTER is a supervised machine learning acoustic event classifier
originally developed for classifying dolphin schools from acoustic
recordings ((Rankin et al. 2017a)). This hierarchical random forest
event classifier relies on an initial call classifier (in the dolphin
case study, this included one each for whistles, echolocation clicks,
and burst pulses) followed by an event classifier. The event classifier
considers the probability of assignment to species for each call in the
call classifier, along with any event-level characteristics (such as
call rate). BANTER is very flexible and can accommodate any number of
measures from any number of detectors. While BANTER relies on multiple
call types, relatively minor changes to a call detector could yield
different detector results ((Rankin et al. 2017b)). In this case,
settings for a whistle \& moan detector were modified to improve
performance at detecting burst pulses. This approach of applying
multiple call detectors (of the same type but with different settings)
suggests that perhaps BANTER could be used on species that only produce
one call type (or where only one call type was analyzed), as long as the
settings of the automated call detector were modified such that the
overall number of detections was different. This approach was
successfully applied to classified echolocation clicks for narwhal and
belugas ((Zahn et al. 2021)). For species who primarily (or exclusively)
produce echolocation clicks, BANTER may serve as an option for automated
machine learning classification.

@@

Recordings from the Eastern North Pacific Ocean off the U.S. west coast
(EPacific) were collected using drifting acoustic recorders during the
2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
survey ((J. Keating et al. 2018a)). Buoys were deployed at a number of
stations in the California Current, including pre-determined locations
(for abundance estimation studies) and in close proximity to seamounts
(Ad Hoc Seamount Experiment). Two types of drifting recorders were
included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New
Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics,
Maynard, MA). Recordings from pre-determined locations were duty cycled
and recorded 2 minutes out of a variable `off' time; recordings from the
Seamount Experiment were continuous. Soundtrap recorders were sampled at
288 kHz and the SM3M recorders were recorded at 256 sample rate (with
some sampling at 96 kHz for ocean noise measurements). Individual buoys
were deployed for 11-23 days (see (J. Keating et al. 2018b) for more
detailed information).

@@

Acoustic detection of beaked whale events were then classified to
species by experienced acoustic researchers using multiple lines of
evidence. If beaked whales were sighted in close proximity to the
recording device, the visual confirmation of species identify was used
to determine species identity. Otherwise, experienced acousticians
followed strict protocol to define species identify based on acoustic
characteristics. When the available characteristics were inconclusive,
the species was considered an unidentified beaked whale; only acoustic
events classified to species were considered for this study. Acoustic
researchers with experience in classifying beaked whale echolocation
clicks examined the bearing-time display for the click detector module
in PAMGuard to assess the species identity of the acoustic event. For
NEFSC data, acoustic detections of beaked whales using towed hydrophone
arrays were further subdivided into click trains based on identification
of consecutive clicks along the same bearing angle in the bearing time
plot (see (DeAngelis et al. 2018b)). Data were saved within the PAMGuard
database and binaries for downstream processing using PAMpal package
((Sakai 2021)) in R programming language ((R Core Team 2022)).

@@

BANTER models was created for each trial using the BANTER package in R
programming language ((Archer 2022a)). Standard click calculations were
used for the BANTER call classifier; event level measures of ICI were
applied to the BANTER event classifier. A minimum of two events per
species and complete suite of variables (absence of NAs) were required
for training and testing of the classification model. The BANTER random
forest models contain two parameters, ntree and sampsize, which must be
selected to ensure stability of the model. The sample size (sampsize) is
the number of events considered for each trial of the model, and the
ntree is the number of trees (iterations of the model) run in a given
model. These values were modified to obtain the best results, with an
emphasis on lower sample size and higher number of trees for
computational performance ((Rankin and Archer 2021)).

The model output include the confusion matrix, and a number of
visualizations provided by the rfpermute package in R programming
language ((Archer 2022b)).

@@

Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed
for this study; species included: Baird's beaked whales (\emph{Berardius
bairdii}), Blainsville's beaked whales (\emph{Mesoplodon densirostris}),
Cuvier's beaked whales (\emph{Ziphius cavirostris}), Cross Seamount
beaked whale (McDonald et al., 2009), Gervais' beaked whales
(\emph{Mesoplodon europaeus}), Longman's beaked whale (\emph{Indopacetus
pacificus}), Sowerby's beaked whales (\emph{Mesoplodon bidens}), True's
beaked whales (\emph{Mesoplodon mirus}), Stejneger's beaked whale
(\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
((Baumann-Pickering et al. 2013a)). Analyses were conducted separately
for the different study areas due to differences in the data collection
methods that precluded combining of datasets. Specific species
combinations and sample sizes varied by study area.

@@

Gervais' and True's beaked whales have similar click characteristics
that make them difficult to differentiate, and require considerable
expertise to classify manually, using a variety of available visuals and
metrics ((DeAngelis et al. 2018c)). Our results suggest that BANTER may
serve as an efficient and effective means of classifying and
differentiating these two species. Despite modest sample sizes, Gervais'
and True's beaked whales showed high classification scores (95.6\% and
100\%, respectively). Unfortunately, we were unable to combine the N and
S Atlantic datasets (which would increase sample sizes of True's beaked
whales) due to differences in hydrophone characteristics that result in
differences in call measures. Calibration of signals (to make them
comparable) may allow for combining datasets to improve sample sizes.

>>>>

\title{Open-Source machine learning BANTER acoustic classification of a
cryptic echolocating species}
\author{Shannon Rankin \and Taiki Sakai \and Frederick Archer \and Jay
Barlow \and Danielle Cholewiak \and Annamaria DeAngelis \and Jennifer L.
K. McCullough \and Erin Oleson \and Melissa Soldevilla}
\date{2/7/23}

\begin{document}
\maketitle
\begin{abstract}
Passive acoustic monitoring is increasingly used for assessing
populations of marine mammals; however, analysis of large datasets is
limited by our ability to easily classify sounds detected.
Classification of beaked whale acoustic events, in particular, require
evaluation of multiple lines of evidence by expert analysts. Here we
present a highly automated approach to acoustic detection and
classification using supervised machine learning and open source
software methods. Data from four large scale surveys of beaked whales
(North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were
analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and
BANTER (hierarchical random forest classifier). Overall correct
classification results ranged from 86\% for the South Atlantic data to
99\% for the North Atlantic. Results for many species could likely be
improved with increased sample sizes and consideration of alternative
automated detectors. These methods provide a highly automated approach
to acoustic detection and classification using open source methods that
can be readily adopted for species and geographic regions.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

Passive acoustic monitoring (PAM) from has proven to be a valuable tool
for studying populations of marine mammals (Parijs et al. 2009);
however, the value of these studies depends on our ability to identify
the sources of the sounds we are monitoring. Historically, experienced
acousticians manually identify stereotyped sounds that could be reliably
attributed to a given species, based on their spectral or temporal
characteristics (Baumann-Pickering et al. 2013a); (Bittle and Duncan
2013); (Rankin and Barlow 2005); (Soldevilla et al. 2008). The dramatic
increase in recordings make it impossible for experienced acousticians
to manually annotate all data, and there has been an increasing need for
the development of automated classification routines that can provide
accurate determinations of the species responsible for the sounds.

Beaked whales are deep diving marine mammals found in offshore waters;
their long dive intervals and cryptic surfacing behavior make them
difficult to study using typical shipboard visual observation methods
(MacLeod 2018). However, beaked whales make stereotyped echolocation
clicks that have temporal and spatial characteristics that can be used
to differentiate species (Baumann-Pickering et al. 2013b). Detection of
beaked whale signals in large datasets is greatly expanding our
understanding of their population structure and the potential impact of
human activities on these species (Barlow et al. 2022);
(Baumann-Pickering et al. 2014); (Simonis et al. 2020). Unfortunately,
manual classification of beaked whale echolocation clicks requires
analysis by trained observers using a number of visual aids to examine
the call characteristics. Development of automated classification
routines, if accurate, serve to improve the efficiency and decrease the
cost of analyzing large datasets.

BANTER is a supervised machine learning acoustic event classifier
originally developed for classifying dolphin schools from acoustic
recordings (Rankin et al. 2017a). This hierarchical random forest event
classifier relies on an initial call classifier (in the dolphin case
study, this included one each for whistles, echolocation clicks, and
burst pulses) followed by an event classifier. The event classifier
considers the probability of assignment to species for each call in the
call classifier, along with any event-level characteristics (such as
call rate). BANTER is very flexible and can accommodate any number of
measures from any number of detectors. While BANTER relies on multiple
call types, relatively minor changes to a call detector could yield
different detector results (Rankin et al. 2017b). In this case, settings
for a whistle \& moan detector were modified to improve performance at
detecting burst pulses. This approach of applying multiple call
detectors (of the same type but with different settings) suggests that
perhaps BANTER could be used on species that only produce one call type
(or where only one call type was analyzed), as long as the settings of
the automated call detector were modified such that the overall number
of detections was different. This approach was successfully applied to
classified echolocation clicks for narwhal and belugas (Zahn et al.
2021). For species who primarily (or exclusively) produce echolocation
clicks, BANTER may serve as an option for automated machine learning
classification.

@@

Recordings from the Eastern North Pacific Ocean off the U.S. west coast
(EPacific) were collected using drifting acoustic recorders during the
2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
survey (J. Keating et al. 2018a). Buoys were deployed at a number of
stations in the California Current, including pre-determined locations
(for abundance estimation studies) and in close proximity to seamounts
(Ad Hoc Seamount Experiment). Two types of drifting recorders were
included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New
Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics,
Maynard, MA). Recordings from pre-determined locations were duty cycled
and recorded 2 minutes out of a variable `off' time; recordings from the
Seamount Experiment were continuous. Soundtrap recorders were sampled at
288 kHz and the SM3M recorders were recorded at 256 sample rate (with
some sampling at 96 kHz for ocean noise measurements). Individual buoys
were deployed for 11-23 days (see {[}J. Keating et al. (2018b){]} for
more detailed information).

@@

Acoustic detection of beaked whale events were then classified to
species by experienced acoustic researchers using multiple lines of
evidence. If beaked whales were sighted in close proximity to the
recording device, the visual confirmation of species identify was used
to determine species identity. Otherwise, experienced acousticians
followed strict protocol to define species identify based on acoustic
characteristics. When the available characteristics were inconclusive,
the species was considered an unidentified beaked whale; only acoustic
events classified to species were considered for this study. Acoustic
researchers with experience in classifying beaked whale echolocation
clicks examined the bearing-time display for the click detector module
in PAMGuard to assess the species identity of the acoustic event. For
NEFSC data, acoustic detections of beaked whales using towed hydrophone
arrays were further subdivided into click trains based on identification
of consecutive clicks along the same bearing angle in the bearing time
plot (see (DeAngelis et al. 2018b)). Data were saved within the PAMGuard
database and binaries for downstream processing using PAMpal package
(Sakai 2021) in R programming language (R Core Team 2022).

@@

BANTER models was created for each trial using the BANTER package in R
programming language (Archer 2022a). Standard click calculations were
used for the BANTER call classifier; event level measures of ICI were
applied to the BANTER event classifier. A minimum of two events per
species and complete suite of variables (absence of NAs) were required
for training and testing of the classification model. The BANTER random
forest models contain two parameters, ntree and sampsize, which must be
selected to ensure stability of the model. The sample size (sampsize) is
the number of events considered for each trial of the model, and the
ntree is the number of trees (iterations of the model) run in a given
model. These values were modified to obtain the best results, with an
emphasis on lower sample size and higher number of trees for
computational performance (Rankin and Archer 2021).

The model output include the confusion matrix, and a number of
visualizations provided by the rfpermute package in R programming
language (Archer 2022b).

@@

Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed
for this study; species included: Baird's beaked whales (\emph{Berardius
bairdii}), Blainsville's beaked whales (\emph{Mesoplodon densirostris}),
Cuvier's beaked whales (\emph{Ziphius cavirostris}), Cross Seamount
beaked whale (McDonald et al., 2009), Gervais' beaked whales
(\emph{Mesoplodon europaeus}), Longman's beaked whale (\emph{Indopacetus
pacificus}), Sowerby's beaked whales (\emph{Mesoplodon bidens}), True's
beaked whales (\emph{Mesoplodon mirus}), Stejneger's beaked whale
(\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
(Baumann-Pickering et al. 2013a). Analyses were conducted separately for
the different study areas due to differences in the data collection
methods that precluded combining of datasets. Specific species
combinations and sample sizes varied by study area.

@@

Gervais' and True's beaked whales have similar click characteristics
that make them difficult to differentiate, and require considerable
expertise to classify manually, using a variety of available visuals and
metrics (DeAngelis et al. 2018c). Our results suggest that BANTER may
serve as an efficient and effective means of classifying and
differentiating these two species. Despite modest sample sizes, Gervais'
and True's beaked whales showed high classification scores (95.6\% and
100\%, respectively). Unfortunately, we were unable to combine the N and
S Atlantic datasets (which would increase sample sizes of True's beaked
whales) due to differences in hydrophone characteristics that result in
differences in call measures. Calibration of signals (to make them
comparable) may allow for combining datasets to improve sample sizes.

### 2023-03-02 natlantic/beakerBanter_natlantic.Rmd
add epacific analysis
Added updated epacific analysis using CCES

banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
saveRDS(banter_data_ici, file='natlantic_banter_data_ici.rds')
```

>>>>

banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
saveRDS(banter_data_ici, file='natlantic_banter_data_ici.rds')

#save update of Acoustic Study
saveRDS(data, 'natlantic_study.rds')
```

### 2023-03-02 satlantic/beakerBanter_satlantic.Rmd
add epacific analysis
Added updated epacific analysis using CCES

---
title: 'BANTER for Beaked Whales: South Atlantic'
author: "Shannon Rankin"
date: "2023-02-03"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r, assign species ID}
GU1605_BW <- read_excel("GU1605_PG_OfflineEvents_20210809.xlsx", sheet = "AllData")#read in data
#create formatted event/species dataframe 
speciesID <- GU1605_BW[c("Id", "Species", "Database")] %>% 
  mutate("event" = paste(Database, Id, sep=".OE")) %>% 
  rename("species" = "Species") %>% 
  mutate("species" = str_replace_all(species, "'", ""),
         "species" = str_replace(species, "Sowerby", "Sowerbys"),
         "species" = str_replace(species, "Blainville", "Blainvilles"),
         "species" = str_replace(species, "Cuvier", "Cuviers")) %>% 
  select(all_of(c("event", "species")))

#Assign species ID to data
data <- setSpecies(data, method='manual', value = speciesID)
#Note: 130 detections are not given species IDs and will be filtered out at a later step.
```

@@

banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
saveRDS(banter_data_ici, file='satlantic_banter_data_ici.rds')
```

@@

```{r, read banter model, include=FALSE, eval=!freshRun}
banter_model <- readRDS('satlantic_banter_model_t5e3s3_t5e3s4.rds')
```

```{r, banter detector model, eval=freshRun}
banter_model <- initBanterModel(banter_data$events)
banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=5e3, sampsize=4, importance = TRUE)
```

@@

```{r, banter event model}
banter_model <- runBanterModel(banter_model, ntree=5e3, sampsize=4)
summary(banter_model)
```

@@

```{r,save banter model, eval=freshRun}
saveRDS(banter_model, 'satlantic_banter_model_t5e3s3_t5e3s4.rds')
```

@@

```{r, read banter ici model, include=FALSE, eval=!freshRun}
banter_model_ici <- readRDS('satlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
```

```{r, banter ici detector model, eval=freshRun}
banter_model_ici <- initBanterModel(banter_data_ici$events)
banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=10e3, sampsize=2, importance = TRUE)
```

@@

```{r, banter ici event model}
banter_model_ici <- runBanterModel(banter_model_ici, ntree=10e3, sampsize=4)
summary(banter_model_ici)
```

@@

```{r, save banter ici model, eval=freshRun}
saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t10e3s2_t10e3s4.rds')
```

@@

![](banter_model_hawaii_confuseMatrix.png)

@@

```{r, proximity plot, include=TRUE}
png(paste(modelname, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
hawaii_proximityPlot <- plotProximity(banter_model_RF)
dev.off()
hawaii_proximityPlot <- plotProximity(banter_model_RF)

png(paste(modelname_ici, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
dev.off()
ici_hawaii_proximityPlot <- plotProximity(banter_model_ici_RF)
```

@@

```{r, importance heat map, include=TRUE}
png(paste(modelname, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
plotImportance(banter_model_RF, plot.type="heatmap")
dev.off()
hawaii_importance <- plotImportance(banter_model_RF, plot.type="heatmap")

png(paste(modelname_ici, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
plotImportance(banter_model_ici_RF, plot.type="heatmap")
dev.off()
ici_hawaii_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
```

@@

```{r, plot votes, include=TRUE}
png(paste(modelname, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_RF)
dev.off()
hawaii_votes <- plotVotes(banter_model_RF)

png(paste(modelname_ici, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_ici_RF)
dev.off()
ici_hawaii_votes <- plotVotes(banter_model_ici_RF)
```

@@

```{r, class priors, include=TRUE}
hawaii_priors <- classPriors(banter_model_RF, NULL)
```

>>>>

---
title: 'BANTER for Beaked Whales: South Atlantic'
author: "Shannon Rankin"
date: "2023-02-08"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r, assign species ID}
GU1605 <- read_excel("GU1605_PG_OfflineEvents_20210809.xlsx", sheet = "BWonly")#read in data
GU1605 <- filter(GU1605, !eventType == "BRAN")
#create formatted event/species dataframe 
speciesID <- GU1605[c("Id", "Species", "Database")] %>% 
  mutate("event" = paste(Database, Id, sep=".OE")) %>% 
  rename("species" = "Species") %>% 
  mutate("species" = str_replace_all(species, "'", ""),
         "species" = str_replace(species, "Sowerby", "Sowerbys"),
         "species" = str_replace(species, "Blainville", "Blainvilles"),
         "species" = str_replace(species, "Cuvier", "Cuviers")) %>% 
  select(all_of(c("event", "species")))

#Assign species ID to data
data <- setSpecies(data, method='manual', value = speciesID)
#Note: 238 detections are not given species IDs and will be filtered out at a later step.
```

@@

banter_data_ici <- export_banter(data, dropSpecies = 'unid', training=TRUE)
saveRDS(banter_data_ici, file='satlantic_banter_data_ici.rds')

#save update of Acoustic Study
saveRDS(data, 'satlantic_study.rds')
```

@@

```{r, read banter model, include=FALSE, eval=!freshRun}
banter_model <- readRDS('satlantic_banter_model_t1e4s2_t1e5s1.rds')
```

```{r, banter detector model, eval=freshRun}
banter_model <- initBanterModel(banter_data$events)
banter_model <- addBanterDetector(banter_model, banter_data$detectors, ntree=1e4, sampsize=2, importance = TRUE)
```

@@

```{r, banter event model}
banter_model <- runBanterModel(banter_model, ntree=1e5, sampsize=1)
summary(banter_model)
```

@@

```{r,save banter model, eval=freshRun}
saveRDS(banter_model, 'satlantic_banter_model_t1e4s2_t1e5s1.rds')
```

@@

```{r, read banter ici model, include=FALSE, eval=!freshRun}
banter_model_ici <- readRDS('satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
```

```{r, banter ici detector model, eval=freshRun}
banter_model_ici <- initBanterModel(banter_data_ici$events)
banter_model_ici <- addBanterDetector(banter_model_ici, banter_data_ici$detectors, ntree=1e4, sampsize=4, importance = TRUE)
```

@@

```{r, banter ici event model}
banter_model_ici <- runBanterModel(banter_model_ici, ntree=1e5, sampsize=1)
summary(banter_model_ici)
```

@@

```{r, save banter ici model, eval=freshRun}
saveRDS(banter_model_ici, 'satlantic_banter_model_ici_t1e4s4_t1e5s1.rds')
```

@@

![](banter_model_satlantic_confuseMatrix.png)

@@

```{r, proximity plot, include=TRUE}
png(paste(modelname, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
satlantic_proximityPlot <- plotProximity(banter_model_RF)
dev.off()
satlantic_proximityPlot <- plotProximity(banter_model_RF)

png(paste(modelname_ici, "satlantic_proximity.png", sep="_"), width = 20, height = 20, units = 'cm', res = 300)
ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
dev.off()
ici_satlantic_proximityPlot <- plotProximity(banter_model_ici_RF)
```

@@

```{r, importance heat map, include=TRUE}
png(paste(modelname, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
plotImportance(banter_model_RF, plot.type="heatmap")
dev.off()
satlantic_importance <- plotImportance(banter_model_RF, plot.type="heatmap")

png(paste(modelname_ici, "satlantic_importance.png", sep="_"), width = 30, height = 25, units = 'cm', res = 300)
plotImportance(banter_model_ici_RF, plot.type="heatmap")
dev.off()
ici_satlantic_importance <- plotImportance(banter_model_ici_RF, plot.type="heatmap")
```

@@

```{r, plot votes, include=TRUE}
png(paste(modelname, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_RF)
dev.off()
satlantic_votes <- plotVotes(banter_model_RF)

png(paste(modelname_ici, "satlantic_votes.png", sep="_"), width = 20, height = 20, units = 'cm',  res = 300)
plotVotes(banter_model_ici_RF)
dev.off()
ici_satlantic_votes <- plotVotes(banter_model_ici_RF)
```

@@

```{r, class priors, include=TRUE}
satlantic_priors <- classPriors(banter_model_RF, NULL)
```
